---
title: "Cox_model_survival_data"
format: html
editor: visual
---

# I-Training the Cox model on all- pairwise log-ratio data format

```{r}
library(coda4microbiome)
library(survival)
library(phyloseq)
```

Assign the loaded object to a new variable 'phy' and display its summary

load the data:

```{r}
phy <- biom
phy
```

Agglomerate(merge) features that have the same genus then summing their abundance

```{r}
phyg <- tax_glom(phy, taxrank = "Genus" )
```

**remove rare taxa**—specifically, those that are absent (i.e., have zero counts) in **more than 95% of the samples**

--\>focus on taxa that are likely to be biologically significant and provide more reliable results

```{r}
phyg_f = filter_taxa(phyg, function(x) (sum(x == 0)/length(x)) <= 0.95, TRUE)
phyg_f
```

Loading coda4microbiome and preparing input data, survival time, status, and covariates

```{r}
library(coda4microbiome)
# Set data
x = data.frame(t(otu_table(phyg_f)))
time = as.numeric(sample_data(phyg_f)$T1Dweek)
status = as.numeric(sample_data(phyg_f)$T1D)
covars = data.frame(sample_data(phyg_f)$Sex, # Gender
                    sample_data(phyg_f)$Antibiotics) # Antibiotics
```

create train/test split( 70% train 30% test)

```{r}
set.seed(42)
sample_ids <- sample_names(phyg_f)
train_ids <- sample(sample_ids, size = floor(0.7 * length(sample_ids)))
test_ids <- setdiff(sample_ids, train_ids)
```

Subsetting the Phyloseq Object into Training and Testing Datasets

```{r}

phy_train <- prune_samples(train_ids, phyg_f)
phy_test  <- prune_samples(test_ids, phyg_f)
```

Prepare training data

```{r}
x_train <- data.frame(t(otu_table(phy_train))) #microbial abundance features
time_train <- as.numeric(sample_data(phy_train)$T1Dweek) #
status_train <- as.numeric(sample_data(phy_train)$T1D)
covars_train <- data.frame(
  Sex = sample_data(phy_train)$Sex,
  Antibiotics = sample_data(phy_train)$Antibiotics
)
```

train the model

```{r}
set.seed(12345)
Trained_model_1 <- coda_coxnet(
  x = x_train,
  time = time_train,
  status = status_train,
  covar = covars_train,
  lambda = "lambda.min",
  alpha = 0.9,
  nfolds = 5,
  showPlots = TRUE,
  coef_threshold = 0
)
```

Prepare the testing data:

```{r}
x_test <- data.frame(t(otu_table(phy_test)))  # Microbial abundance features
time_test <- as.numeric(sample_data(phy_test)$T1Dweek)
status_test <- as.numeric(sample_data(phy_test)$T1D)
covars_test <- data.frame(
  Sex = sample_data(phy_test)$Sex,
  Antibiotics = sample_data(phy_test)$Antibiotics
)
```

```{r}
coef_min <- coef(Trained_model_1$fit, s = Trained_model$lambda_min)

```

prepare testing data and generate the risk scores

```{r}

# Step 1: Extract and clean OTU table
otu_mat <- as.data.frame(t(otu_table(phy_test)))

colnames(otu_mat) <- make.names(colnames(otu_mat))  
# Step 2: Extract coefficients and assign names
coeffs <- Trained_model_1$`log-contrast coefficients`
taxa_names <- Trained_model_1$taxa.name
names(coeffs) <- taxa_names  # Now coeffs has names
# Step3: select specific taxa
selected_taxa <- names(coeffs)  # Now this is NOT NULL
#subset the coloumns listed in the selected taxa
x_test_selected <- otu_mat[, selected_taxa, drop = FALSE]
#x_test_log <- x_test_selected #log(x_test_selected + 1e-6)
# Step 4: Compute risk scores
risk_score_test <- as.vector(as.matrix(x_test_selected) %*% coeffs)
# Use the correct variable name (matching case)
#x_test <- data.frame(t(otu_table(phy_test)))[                                  #Trained_model$selected_features, drop = FALSE] #non existed

# Extract survival time from sample data
time_test <- as.numeric(sample_data(phy_test)$T1Dweek)

# Extract survival status (change 'status' to your actual column name if necessary)
status_test <- as.numeric(sample_data(phy_test)$T1D) #"status columns not exists, T1D instead

# Make sure covar_names is defined (same covariates used in training)
covar_names <- colnames(covars_train)

# Extract covariates from the test data (matching covar_names)
covars_test <- data.frame(sample_data(phy_test))[ , covar_names, drop = FALSE]

```

evaluate the performance

```{r}

#this code is not working because the predict need a survival object!
#See step above to generate the individual risk score from the test data
c_index <- survConcordance(Surv(time_test, status_test) ~ risk_score_test)$concordance
print(paste("Test set C-index:", round(c_index, 3)))
```

survival curve plot:

```{r}
survcurve <- plot_survcurves(risk.score = risk_score_test,
                             time_test,
                             status_test,
                             strata.quantile = 0.5)
survcurve


T1Dmodel
```
