---
title: "Cox_model_survival.qmd"
format: html
editor: visual
---

## Evaluating the Impact of Data Transformation Formats on Cox Proportional Hazards Model Performance

```{r}
library(phyloseq)
```

Load the dataset

```{r}
load("biom.Rdata")
```

Assign the loaded object to a new variable 'phy' and display its summary

```{r}
phy <- biom
phy
```

agglomerate all the features of the same genus in a single variable

```{r}
phyg <- tax_glom(phy, taxrank = "Genus" )
phyg
```

Filter out taxa that are absent (i.e., zero) in more than 95% of the samples

```{r}
phyg_f = filter_taxa(phyg, function(x) (sum(x == 0)/length(x)) <= 0.95, TRUE)
phyg_f
```

Loading coda4microbiome and preparing input data, survival time, status, and covariates

```{r}
library(coda4microbiome)
# Set data
x = data.frame(t(otu_table(phyg_f)))#extracts the OTU matrix from phyloseq object and transform it to a dataframe
time = as.numeric(sample_data(phyg_f)$T1Dweek)
status = as.numeric(sample_data(phyg_f)$T1D)
covars = data.frame(sample_data(phyg_f)$Sex, # Gender
                    sample_data(phyg_f)$Antibiotics) # Antibiotics
```

I-Fitting a penalized Cox regression model into logratio data format

```{r}
set.seed(12345)
T1Dmodel <- coda_coxnet(as.matrix(x),#dataframe
                        time,#time to event
                        status,#event occurrence 0/1
                        covar=covars, #data frame with covariates which are variables that might influence the time to event,
                        lambda = "lambda.1se",#penalization parameter,controls the strength of the penalty applied to the model, controls the overfitting
                        alpha = 0.9, nfolds = 5,#elatic net parameter
                        showPlots = FALSE,#no plots
                        coef_threshold = 0.01)#minimum absolute value of the coefficient for a variable to be included in the model
T1Dmodel

survcurve <- plot_survcurves(risk.score = T1Dmodel$risk.score,
                             time,
                             status,
                             strata.quantile = 0.5)
survcurve
```

Plot analysis:

.based on the **Microbial risk plot**: higher presence of T1D develop- ment events for those individuals with higher microbial risk scores.

.based on the **Microbial signature for T1D onset plot**, X262095 and X4407703 are the most significant taxonomy that influence the development of type 1 diabetes / 10 taxa with positive coefficient and 11 taxa with negative coefficient.

.based on the **survival curve**: aim to assess the possible association between the microbial risk score and the time to development of T1D. red line: individuals with higher microbial risk score shows shorter times of T1D. blue line:individuals presenting lower values of microbial risk scores develops T1D at later time.

II- ilr Transformation and Cox Model Implementation

```{r}
library(survival)
set.seed(12345)
T2Dmodel <- abund_coxnet2(phyg_f,#dataframe
                        time,#time to event
                        status,#event occurrence 0/1
                        covar=covars, #data frame with covariates which are variables that might influence the time to event,
                        lambda = "lambda.1se",#penalization parameter,controls the strength of the penalty applied to the model, controls the overfitting
                        alpha = 0.9, nfolds = 5,#elatic net parameter
                        showPlots = FALSE,#no plots
                        coef_threshold = 0.01, transform = "ilr")#minimum absolute value of the coefficient for a variable to be included in the model


# Extracting selected taxa and log-contrast coefficients from the fitted T1D model
T2Dmodel$taxa.num
T2Dmodel$taxa.name

#risk score
T2Dmodel$`risk_score_plot`
#signature plot
T2Dmodel$`signature_plot`
#survival curve
survcurve_2 <- plot_survcurves(risk.score = T2Dmodel$risk.score,
                             time,
                             status,
                             strata.quantile = 0.5)
survcurve_2
```

.based on the **Microbial signature for T1D onset plot**, X262095 and X177271 are the most significant taxonomy that influence the development of type 1 diabetes / 10 taxa with positive coefficient and 9 taxa with negative coefficient.

III- alr Transformation and Cox Model Implementation

```{r}
set.seed(12345)
T3Dmodel <- abund_coxnet2(phyg_f,#dataframe
                          time,#time to event
                          status,#event occurrence 0/1
                          covar=covars, #data frame with covariates which are variables that might influence the time to event,
                          lambda = "lambda.1se",#penalization parameter,controls the strength of the penalty applied to the model, controls the overfitting
                          alpha = 0.9, nfolds = 5,#elatic net parameter
                          showPlots = FALSE,#no plots
                          coef_threshold = 0.01, transform = "alr")#minimum absolute value of the coefficient for a variable to be included in the model

# Extracting selected taxa and log-contrast coefficients from the fitted T1D model
T3Dmodel$taxa.num
T3Dmodel$taxa.name

#risk score
T3Dmodel$`risk_score_plot`
#signature plot
T3Dmodel$`signature_plot`
#survival curve
survcurve_3 <- plot_survcurves(risk.score = T3Dmodel$risk.score,
                             time,
                             status,
                             strata.quantile = 0.5)
survcurve_3
```

.based on the **Microbial signature for T1D onset plot**, X262095 and X177271 are the most significant taxonomy that influence the development of type 1 diabetes / 10 taxa with positive coefficient and 9 taxa with negative coefficient.

IV-Relative abundance Transformation and Cox Model Implementation

```{r}
set.seed(12345)
T4Dmodel <- abund_coxnet2(phyg_f,#dataframe
                          time,#time to event
                          status,#event occurrence 0/1
                          covar=covars, #data frame with covariates which are variables that might influence the time to event,
                          lambda = "lambda.1se",#penalization parameter,controls the strength of the penalty applied to the model, controls the overfitting
                          alpha = 0.9, nfolds = 5,#elatic net parameter
                          showPlots = FALSE,#no plots
                          coef_threshold = 0.01, transform = "rel")#minimum absolute value of the coefficient for a variable to be included in the model

# Extracting selected taxa and log-contrast coefficients from the fitted T1D model
T4Dmodel$taxa.num
T4Dmodel$taxa.name

#risk score
T4Dmodel$`risk_score_plot`
#signature plot
T4Dmodel$`signature_plot`
#survival curve
survcurve_4 <- plot_survcurves(risk.score = T4Dmodel$risk.score,
                             time,
                             status,
                             strata.quantile = 0.5)
survcurve_4
```

based on the **Microbial signature for T1D onset plot** X138389 is the only taxa that can contribute to the development of T1D.

V-clr Transformation and Cox Model Implementation

```{r}
set.seed(12345)
T5Dmodel <- abund_coxnet2(phyg_f,#dataframe
                          time,#time to event
                          status,#event occurrence 0/1
                          covar=covars, #data frame with covariates which are variables that might influence the time to event,
                          lambda = "lambda.1se",#penalization parameter,controls the strength of the penalty applied to the model, controls the overfitting
                          alpha = 0.9, nfolds = 5,#elatic net parameter
                          showPlots = FALSE,#no plots
                          coef_threshold = 0.01, transform = "clr")#minimum absolute value of the coefficient for a variable to be included in the model

# Extracting selected taxa and log-contrast coefficients from the fitted T1D model
T5Dmodel$taxa.num
T5Dmodel$taxa.name

#risk score
T5Dmodel$`risk_score_plot`
#signature plot
T5Dmodel$`signature_plot`
#survival curve
survcurve_5 <- plot_survcurves(risk.score = T5Dmodel$risk.score,
                             time,
                             status,
                             strata.quantile = 0.5)
survcurve_5
```

.based on the **Microbial signature for T1D onset plot**, X262095 and X4407703 are the most significant taxonomy that influence the development of type 1 diabetes / 9 taxa with positive coefficient and taxa with negative coefficient.

VI-no transformation

```{r}
set.seed(12345)
T6Dmodel <- abund_coxnet2(phyg_f,#dataframe
                          time,#time to event
                          status,#event occurrence 0/1
                          covar=covars, #data frame with covariates which are variables that might influence the time to event,
                          lambda = "lambda.1se",#penalization parameter,controls the strength of the penalty applied to the model, controls the overfitting
                          alpha = 0.9, nfolds = 5,#elatic net parameter
                          showPlots = FALSE,#no plots
                          coef_threshold = 0.01, transform = "none")#minimum absolute value of the coefficient for a variable to be included in the model

# Extracting selected taxa and log-contrast coefficients from the fitted T1D model
T6Dmodel$taxa.num
T6Dmodel$taxa.name

#risk score
T6Dmodel$`risk_score_plot`
#signature plot
T6Dmodel$`signature_plot`
#survival curve
survcurve_6 <- plot_survcurves(risk.score = T6Dmodel$risk.score,
                             time,
                             status,
                             strata.quantile = 0.5)
survcurve_6
```
