---
title: "TreeSE_Coxmodel.qmd"
format: html
editor: visual
---

## This report focuses on evaluating the predictive performance of the Cox proportional hazards model under transformed and un-transformed data. The primary objective is to identify which transformation method allows the Cox model to yield the most accurate survival predictions. By systematically applying distinct transformations to the data-set and assessing the model’s performance, this study aims to provide insights into optimizing data pre-processing for survival analysis.

## I-Fitting a Cox proportional hazard model into all- pairwise log-ratio data format

Load libraries

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(mia)
library(ggplot2)
library(lattice)
library(TreeSummarizedExperiment)
library(coda4microbiome)
library(SummarizedExperiment)
library(tibble)
library(phyloseq)
library(survival)
library(survival)
```

link to download the biom.Rdata dataset: https://github.com/wg99526/MiSurvGit/tree/main/Data

```{r}
load("~/Desktop/Microbiome_Analysis/TreeSE/biom.Rdata")
```

convert phyloseq object to TreeSE

```{r}
tse <- convertFromPhyloseq(biom)
```

split the data into training and testing subsets

```{r}
# Set seed for reproducibility
set.seed(123)
# Split TreeSE into training and test sets by samples (columns)
num_samples <- ncol(tse)
train_indices <- sample(seq_len(num_samples), size = floor(0.7 * num_samples))
test_indices <- setdiff(seq_len(num_samples), train_indices)
# Subset TreeSE
tse_train <- tse[, train_indices]
tse_test <- tse[, test_indices]
```

Prepare training data

```{r}
# Extract colData
col_data_train <- as.data.frame(colData(tse_train))
# Extract time-to-event and status from colData
time <- col_data_train$T1Dweek   # numeric vector
status <- col_data_train$T1D  # binary vector (0 = censored, 1 = event)
covars <- col_data_train[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
x <- t(assay(tse_train))  # transpose: rows = samples, columns = features
# Replace row names with only digits extracted from original row names
rownames(x) <- gsub("\\D", "", rownames(x))
colnames(x) <- gsub("\\D", "", colnames(x))
```

Prepare testing data

```{r}
#prepare the testing subset
col_data_test <- as.data.frame(colData(tse_test))
# Extract time-to-event and status from colData
time_test <- col_data_test$T1Dweek   # numeric vector
status_test <- col_data_test$T1D  # binary vector (0 = censored, 1 = event)
covars_test <- col_data_test[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
x_test <- t(assay(tse_test))  # transpose: rows = samples, columns = features
# Replace row names with only digits extracted from original row names
rownames(x_test) <- gsub("\\D", "", rownames(x_test))
colnames(x_test) <- gsub("\\D", "", colnames(x_test))
```

transform the assay to all pairwise log ratio format

```{r}
coef_threshold = 0
lambda = "lambda.1se" 
nvar = NULL
alpha = 0.9
nfolds = 10
showPlots = TRUE
x <- impute_zeros(x)
kselect <- ncol(x)
taxaselect <- (1:ncol(x))
lrmatrix <- logratios_matrix(x)
lrX <- lrmatrix[[1]]
idlrX <- lrmatrix[[2]]
nameslrX <- lrmatrix[[3]]
#transform the assay from the testing subset:
lrmatrix_test <- logratios_matrix(x_test)
lrX_test <- lrmatrix[[1]]
```

implement cox model

```{r}
set.seed(123)
y = Surv(time, status)# Cox response variable (time & status)
# fits a standard Cox proportional hazards model (coxph) using all covariates
df0 <- data.frame(as.matrix(y), covars)
model0 <- coxph(Surv(time, status) ~ ., data = df0)
x0 <- predict(model0)#linear predictor
#Penalized Cox Model with Offset
cvfit <- glmnet::cv.glmnet(lrX, y, family = "cox",
                               type.measure = "C",
                               nfolds = nfolds, alpha = alpha,
                               keep = TRUE,
                               offset = x0)
plot(cvfit)
predictions<-as.numeric(predict(cvfit,lrX,s=lambda, newoffset=x0))
Cindex_signature <- glmnet::Cindex(pred = predictions, y)
```

Generate plots

```{r}
rowlasso <- max(which(cvfit$glmnet.fit$df <= nvar))
lambdavalue <- lambda
  if (is.character(lambda)) {
    if (lambda == "lambda.1se")
      lambdavalue <- cvfit$lambda.1se
    if (lambda == "lambda.min")
      lambdavalue <- cvfit$lambda.min
  }
idrow <- max(which(cvfit$glmnet.fit$lambda >= lambdavalue))
coeflr <- as.vector(coef(cvfit, s = lambda)) #[-1]
lrselect <- which(coeflr != 0)
coeflogcontrast <- rep(0, ncol(x))
for (i in (1:length(coeflr))) {
coeflogcontrast[idlrX[i, 1]] <- coeflogcontrast[idlrX[i,1]] +  coeflr[i]
coeflogcontrast[idlrX[i, 2]] <- coeflogcontrast[idlrX[i,2]] -   coeflr[i]
  }
varlogcontrast <- which(abs(coeflogcontrast) > coef_threshold)
coeflogcontrast <- coeflogcontrast[varlogcontrast]

(names.select <- colnames(x)[varlogcontrast])
(sign <- ifelse(coeflogcontrast > 0, 1, 0))
sign <- factor(sign, levels = c(0, 1), labels = c("negative",
                                                    "positive"))
logcontrast = as.matrix(log(x)[, varlogcontrast]) %*% coeflogcontrast # Bacterial signature



#predictions<-x0+logcontrast
predictions<-as.numeric(predict(cvfit,lrX,s=lambdavalue, newoffset=x0))

coeflogcontrast<-2*coeflogcontrast/sum(abs(coeflogcontrast))

if (length(varlogcontrast) == 0){
    Cindex_signature <- 0.5
  } else {
    Cindex_signature <- glmnet::Cindex(pred=predictions, y)     # Apparent C-Index
  }
mcvCindex <- cvfit$cvm[idrow]
sdcvCindex <- cvfit$cvsd[idrow]

plot1<-NULL
plot2<-NULL

if (length(lrselect>0)){

plot1 <- plot_riskscore(predictions, x, time, status, showPlots = showPlots)
plot2 <- plot_signature(names.select, coeflogcontrast, showPlots = showPlots)

  } else {
    print("No variables are selected. The risk score plot and the signature plot are not displayed.")
  }
```

## II-Fitting a Cox proportional hazard model into centered log-ratio data format

```{r}
tse <- convertFromPhyloseq(biom)
```

implement cox model on un-transformed data

```{r}
set.seed(123)
y = Surv(time, status)# Cox response variable (time & status)
# fits a standard Cox proportional hazards model (coxph) using all covariates
df0 <- data.frame(as.matrix(y), covars)
model1 <- coxph(Surv(time, status) ~ ., data = df0)
x0 <- predict(model1)#linear predictor
#Penalized Cox Model with Offset
cvfit1 <- glmnet::cv.glmnet(x, y, family = "cox",
                               type.measure = "C",
                               nfolds = nfolds, alpha = alpha,
                               keep = TRUE,
                               offset = x0)
plot(cvfit1)
predictions1<-as.numeric(predict(cvfit1,x,s=lambda, newoffset=x0))
Cindex_signature <- glmnet::Cindex(pred = predictions1, y)
```

Generate plots

```{r}
rowlasso <- max(which(cvfit$glmnet.fit$df <= nvar))

lambdavalue <- lambda
if (is.character(lambda)) {
  if (lambda == "lambda.1se")
    lambdavalue <- cvfit$lambda.1se
  if (lambda == "lambda.min")
    lambdavalue <- cvfit$lambda.min
}

idrow <- max(which(cvfit$glmnet.fit$lambda >= lambdavalue))
coeflr <- as.vector(coef(cvfit, s = lambda))
lrselect <- which(coeflr != 0)

coeflogcontrast <- rep(0, ncol(x))
for (i in (1:length(coeflr))) {
  coeflogcontrast[idlrX[i, 1]] <- coeflogcontrast[idlrX[i, 1]] + coeflr[i]
  coeflogcontrast[idlrX[i, 2]] <- coeflogcontrast[idlrX[i, 2]] - coeflr[i]
}

varlogcontrast <- which(abs(coeflogcontrast) > coef_threshold)
coeflogcontrast <- coeflogcontrast[varlogcontrast]

names.select <- colnames(x)[varlogcontrast]
sign <- ifelse(coeflogcontrast > 0, 1, 0)
sign <- factor(sign, levels = c(0, 1), labels = c("negative", "positive"))

# Removed logcontrast line (log transformation)

predictions <- as.numeric(predict(cvfit1, x, s = lambdavalue, newoffset = x0))



  Cindex_signature <- glmnet::Cindex(pred = predictions, y)


mcvCindex <- cvfit$cvm[idrow]
sdcvCindex <- cvfit$cvsd[idrow]

plot1 <- NULL
plot2 <- NULL

if (length(lrselect > 0)) {
  plot1 <- plot_riskscore(predictions, x, time, status, showPlots = showPlots)
  plot2 <- plot_signature(names.select, coeflogcontrast, showPlots = showPlots)
} else {
  print("No variables are selected. The risk score plot and the signature plot are not displayed.")
}

```

evaluate model performance on transformed and un-transformed data

```{r}
set.seed(123)
# Define models and their corresponding test matrices
models <- list(
transformed = list(model = model0, cvfit = cvfit, x_test = lrX_test, x_train = lrX),
untransformed = list(model = model1, cvfit = cvfit1, x_test = x_test, x_train = x)
)

# Prepare result list to store C-index for each version
cindex_results <- list()#create an empty list

# Loop through each model type
for (type in names(models)) {
cat("\n--- Testing:", type, "---\n")
  
# Extract components
current_model <- models[[type]]$model
current_cvfit <- models[[type]]$cvfit
current_x_test <- models[[type]]$x_test
  
# Ensure consistent row filtering
complete_cases <- complete.cases(covars_test) & 
                  complete.cases(current_x_test) & 
                  !is.na(time_test) & !is.na(status_test)
  
# Clean testing data
covars_test_clean <- covars_test[complete_cases, , drop = FALSE]
x_test_clean <- current_x_test[complete_cases, , drop = FALSE]
time_test_clean <- time_test[complete_cases]
status_test_clean <- status_test[complete_cases]
  
# Baseline prediction using covariates only
x0_test_clean <- predict(current_model, 
                           newdata = as.data.frame(covars_test_clean), 
                           type = "lp")
  
# Predict using penalized model
lambda_value <- current_cvfit$lambda.min
test_predictions <- predict(current_cvfit, 
                              newx = x_test_clean, 
                              s = lambda_value, 
                              newoffset = x0_test_clean, 
                              type = "link")
  
# Calculate C-index
test_cindex <- glmnet::Cindex(pred = as.numeric(test_predictions),
y = Surv(time_test_clean, status_test_clean))
  
# Store results
cindex_results[[type]] <- test_cindex
  
# Print for inspection
cat("Samples used:", length(time_test_clean), "\n")
cat("C-index:", round(test_cindex, 4), "\n")
}
```
